Blocks of extended ancestry around selected loci in hybrid zones

### 2016-02-29 drafts/hybrid_zone.tex
alisa's edits

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity to better understand the mechanisms underlying reproductive isolation and the process of speciation. \yb{Some form of selection against hybrid genotypes is likely common in many hybrid zones. Here, we use a combination of analytical theory and explicit forward simulations} to gain a better understanding of how this impacts patterns of introgression \yb{across genomic and geographic} space. \alisa{We describe the properties of clines across the genomes as they are formed, and find that locally disadvantageous alleles exist as recent, small families that move relatively rapidly across space}. \yb{We find that, within a population, ancestral blocks surrounding single-locus incompatibilities} are longer than average \yb{ancestry blocks} within the genome, especially when removed from the center of the hybrid zone.  The findings presented here suggest a means for characterizing hybrid zones and identifying candidate targets of selection.
\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}

@@

\plr{
Main points, make sure they appear in the abstract:
clines establish over time $1/s$, extend along genome to at least $1/T$;
locally disadvantageous alleles exist as recent, small families (moving at speed $\sigma/\sqrt{s}$);
haplotype chunks longest around introgressing selected bits, flanked by shorter chunks.
}

@@

The process of speciation commonly involves allopatric divergence in relative isolation \citep{Coyne2004}.  
When formerly isolated populations come back into contact 
before reproductive isolation is complete, gene flow is still possible. 
As two such populations interbreed,
individuals' genomes can be labeled according to which ancestral population they are inherited from,
and local interbreeding and migration of such populations
creates a gradient of ancestry assignments across geography,
centered on their point of contact \citep{Barton1985}. 
%\plr{(check that neutral clines discussed in barton '85)}
If the populations are sufficiently diverged,
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
leading to substantial correlations in ancestry between distant markers.

Gene flow will eventually erase 
both geographic clines in ancestry, and strong correlations in ancestry.
Subsets of the genome can maintain stable clines in ancestry if selection against migrant, or hybrid, genotypes impedes gene flow across the hybrid zone \citep{Barton1979a}. The influence of selection on linked loci also prevents the mixing of ancestry at regions surrounding these loci, the effect of which decreases with physical genomic distance \citep{Barton1986,Barton1983}. 
The mixing of ancestries of a given genome is mediated by 
the process of recombination, forming chromosomal \emph{junctions} in ancestry \citep{Fisher1954, Chapman2002, Baird2003}.   
The expected distribution of lengths of unbroken ancestry in admixed populations has been previously described under neutral conditions both in single populations and in contact zones \citep[e.g.][]{Gravel2012,Hellenthal2014,Sedghifar2015}.

% Variation in \yb{ the action of } selection against migrant alleles within the genome results in genome-wide heterogeneity in cline width, as the frequency of foreign ancestry surrounding targets of selection is reduced \citep{Barton1979}. 
% While these patterns of co-ancestry have been described in multi-locus systems, the \yb{expected length of ancestral tracts around selected loci in admixing populations remains poorly understood}\alisa{, especially in the geographic context of hybrid zones.}
\yb{While empirical patterns of the block length distribution in hybrid zones have rarely been used to inform the history of selection,}  \alisa{the availability of dense genotype data in combination with linkage information pave a way to utilize such patterns to better understand the processes operating in hybrid zones. } %GREAT SENTENCE!!
Here, we investigate expected patterns of co-ancestry \yb{ --- that is, the genomic scale at which ancestry strongly covaries between sites --- } across a hybrid zone in the presence of selection.

The extensive history of population genetic modeling of hybrid zones has focused largely on the equilibrium states of ancestry clines \citep{Barton1979a,Barton1986}.  
Since stronger selection reduces the amount of genetic mixing,
these models predict the narrowest clines 
(i.e., the most rapid changes in allele frequencies)
at loci close to selected loci, and wider clines at less tightly linked loci.
This provides some expectation for how selection shapes covariation in ancestry at a single locus over \emph{geographic space}.
Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry over \emph{genomic space}.

Covariance in ancestry across a genome provides information about the number of surviving recombination events that have taken place between ancestral genotypes and therefore provides information about the age of the hybrid zone and the relative strength of selection affecting subsets of the genome. Our goal is to understand how selection against hybrids interacts with dispersal and recombination to produce observed patterns ancestry around selected sites in recently established hybrid zones. In addition to interpreting non-equilibrium patterns of diversity in hybrid zones, this could be especially important for understanding the impact of selection on the extent of mixing of ancestral genomes and, consequently, the maintenance of species distinctness. In the extreme case of zero fitness in first-generation hybrids, for example, parental genomes will never mix and the stable hybrid zone will resemble a recently formed one. 
If, on the other hand, hybrids have non-zero fitness, some amount of introgression can occur, and  patterns of introgression can provide information about selection \yb{in addition to the} extent of hybridization, and timing of secondary contact \citep{Price2009, Hellenthal2014}. 
\yb{In this latter case, heterogeneous patterns of ancestry across the genome potentially allows researchers to identify putative targets of selection in hybridizing populations.}  
\yb{To date, researchers have done so by examining heterogeneous clines in ancestry across loci \citep{Porter1997, Gompert2012}, but have not made use of the information contained in the length of ancestry blocks around putatively selected loci.}

Under a model of selection against hybrids, migrant haplotypes containing the selected locus will be removed rapidly from the population, preventing introgression in surrounding genomic regions. We therefore expect \yb{a deficit of} short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. As a corollary to this, conditional on having the ancestry that is at lower frequency (that is, being on the `wrong' side of the hybrid zone), the length of unbroken ancestry surrounding the selected locus will be relatively long when far away from the zone center.  This is because an unfit haplotype is more likely to be a recent migrant from the other side of the hybrid zone, and therefore will not have experienced many recombination events.

Using a combination of predictions from theory and simulated hybrid zones, we present a model of genome-wide patterns coancestry as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. For this, we build on the neutral framework presented in \citet{Sedghifar2015} to derive the expected distribution of ancestry block lengths across a contact zone.  
\yb{The theoretical expectations developed in this manuscript can aid in the identification of loci under selection in hybridizing populations.}

@@

We model selection through only a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations,
and individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are effectively indistinguishable from models of ecological selection or epistatic systems \cite{Kruuk1999,Barton2000a}.

As in \citet{Sedghifar2015}, 
we gain substantial insight by considering where the ancestors of modern-day individuals lived,
i.e., where their lineages fall across geography as we look further back towards the time of first contact.
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$ at the time of secondary contact,
i.e., if its lineage traces back to to the $A$ side.
A block of genome is of ancestry $A$ if every locus in it does the same;
recombination events may cause us to follow more than one lineage back to determine the block's ancestry,
and the block will be of ancestry $A$ if all lineages trace back to the $A$ side.
This process, in which recombination events cause such lineages to \emph{branch},
is depicted in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if the block includes the selected site then it is necessarily on the $A$ background.
As the identity of the selected allele determines how selection acts on the haplotype,
and linked variation can only move between the $A$ and the $B$ background in heterozygotes,
a key factor in these models is the density and fecundity of heterozygotes.

@@

\paragraph{Establishment of the cline at the selected locus}
The deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citep{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, even in steady state, 
a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on numerous approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. 
%\alisa{I don't think I understand the independence from $\sigma$?}
%\plr{It gets set up at speed $\sigma$ over a distance proportional to $\sigma$.}

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele $p(x,t)$ is known.
We will understand first how lineages of sampled individuals behave at the selected locus,
from which lineages at linked loci require only a minor modification.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $\sigma^2 \grad \log p(x,t)$ (i.e., it moves at speed $\sigma^2$ in the potential $-\log p$).
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x+r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
\plr{Does that help at all?}

At a geographic position far from the center, $p=1$ to the left, and is proportional to  $\exp(-x\sqrt{s}/\sigma)$ to the right.
Therefore, roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma/\sqrt{s}$ 
back towards the side where they are more common. 
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact, this push must get more intense the closer it is to the time of secondary contact.
Lineages of $B$ alleles do the same thing, of course, but in the opposite direction. 
% \alisa{I want to make a figure from the simulations that show the rapid retreat of wrong ancestries back to where they came from}

@@

Suppose we sample an allele today at a locus at recombination distance $r$ to the selected site.
We will follow its lineage back through time, using $\tau$ to denote ``time ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ (same as previously described movement of a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$)or $r p$ (for $B$ to $A$).

@@

\alisa{This process of a haplotype sampled in the present day tracing it's path $X_\tau$ backwards in time, while splitting and changing backgrounds $Z_\tau$ through recombination, is shown in Figure~\ref{Fig:schematic}.}
\plr{Connect these up more.}

@@

\paragraph{Numerical solutions}
Notice that equations \eqref{eqn:g_pde}
are hierarchical in $[a,b]$:
the equation for haplotype identity probabilities on a segment $[a,b]$ depends only on those probabilities for segments contained in $[a,b]$.
This allows for efficient numerical solutions,
described in more detail in Appendix \ref{apx:haplotype_calcs}.

@@

\begin{figure}
\includegraphics{figs/alleleFrequencies_sim}
\caption{
    Frequency of ancestry B,
    across geography and at several physical positions on the genome, simulated for a hybrid zone
     $T=100$ generations after secondary contact,
    and with $s=0.1$.
        50 demes, each with a population size of 500 diploid individuals, were simulated across the hybrid zone.
    Each line represents a locus some distance $r$ away from the true target of selection, 
    and $r=0$ represents the locus that is under selection. Grey lines represent the same positions from a simulation with
    identical parameters except $s=0$.
    Corresponding theoretical quantities are shown juxtaposed in Figure \ref{sfig:alleleFreq_tau100_comparison}.
}\label{alleleFreq_tau100}
\end{figure}

@@

The above theory and simulations allow us to make predictions about patterns surrounding selected loci. In reality, however, such loci are not known, so it is useful to have per-site statistics that may allow for detection of candidate targets of selection. The most straightforward measure is $l_I(m)$,  the mean length of all haplotype chunks that contain genomic position $m$, given that $m$ is of ancestry $I$, where $I\in\{A,B\}$.
We then focus on $I_B(m)$ for $m$ across a chromosome. 
%Another statistic is the cumulative probability $C_I(m)$ of a sampled chunk encompassing position $m$ being longer than a chunk sampled from the distribution of all chunk lengths across the genome in the population.

@@

Although our analytical model does not account for coalescence, we are able to partially trace the genealogy of haplotypes in our simulations. In particular, we wish to know the relative number of ancestors from $T$ generations ago that are represented in present day populations at a given locus, given the frequency of a particular ancestry. This is a reflection of the average size of a haplotype family, and is potentially a source of additional information. Within each deme $x$, we calculated $F_B(m,x)$, which is the number of independent ancestors from time $T$ contributing to the pool of alleles of ancestry $B$ at site $m$, divided by the number of individuals of ancestry $B$ at site $m$.
%\plr{how about $x$ instead of $X$? above $x$ refers to spatial location.}

@@

\paragraph{The timescale over which linked clines relax:}
Although loci not under selection can in principle spread across the cline unimpeded,
in practice it can take quite some time for even unlinked neutral loci to homogenize,
due to the decreased fitness of heterozygotes \citep{Barton1986}
and the relative slowness of diffusive movement \citep{Sedghifar2015}.
We are interested, for samples across a cline $T$ units of time after secondary contact, how far away from the selected locus one has to look to see clines that resemble the genome-wide mean.

@@

\plr{Figure showing this? 
    For instance, either: (a) time-genome heatmap of number of heterozygotes (from PDE);
    or (b) a few time-space heatmaps of frequencies at different values of r (from PDE);
    or (c) a few space-genome heatmaps of frequencies at different times (from sims)
}

@@

@@

% This pattern provides a potential avenue for identifying targets of selection by looking for a signal of increased $l_B$ or $l_A$ for positions surrounding a selected site, compared to the genome-wide background  at geographic sites far from the center of the zone (Figure~\ref{Fig:blockLengths},~\ref{Supp:blockLengthHeatmap}). 
The mean haplotype length found without conditioning on ancestry ($l_\cdot$)
also shows an small increase near the selected locus,
but it is much smaller (see Figures \ref{Supp:blockLengthHeatmapNoAnc} and \ref{Supp:blockLengthNoAnc}).

\alisa{For very strong $s$, very few haplotypes are found on the ``wrong" side, and so the signal of longer haplotypes around selected loci may only be present at geographic locations close to the center of the zone. The overall length of the haplotypes decreases as the hybrid zone ages, but the signal remains strong relative to the rest of the genome and detectability will largely depend on the density of ancestry-informative markers.}

@@

We expect that, in the absence of selection, blocks of $A$ ancestry across the genome 
will tend to be shorter the further one goes into the $B$ side of the cline,
because they have had more opportunities to recombine with $B$ haplotypes. 
However, we expect that stretches of $A$ ancestry containing a selected site will be longer
than those that do not contain the selected site at the same spatial location,
because lineages containing the selected site have usually been inherited from the $A$ side of the cline recently. 
As discussed above, we expect these lineages move at speed roughly $\sigma/\sqrt{s}$,
so (selected) $A$ alleles at distance $x$ from the cline center
to have last had an ancestor on the $A$ side of the cline around $x \sqrt{s}/\sigma$ generations ago 
(compared to $x^2/\sigma^2$ for a neutral allele),
and so the enclosing $A$ haplotype is no longer than $\sigma/(x \sqrt{s})$.
%\plr{(check predicted decay of mean length $1/x$ in simulations?)}
For linked haplotypes not containing the selected site, the situation is more complicated, and harder to intuit.

@@

\alisa{These seem like an orphan statement. Any idea where to stick these?}
We further find that the long blocks of ancestry around selected sites on the wrong side of the hybrid zone will be descended from a relatively large number of independent ancestors, per individual, of the less fit ancestry. This result reflects that, on average, less fit haplotypes tend to be recent migrants with few descendants.

>>>>

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity to better understand the mechanisms underlying reproductive isolation and the process of speciation. \yb{Some form of selection against hybrid genotypes is likely common in many hybrid zones. Here, we use a combination of analytical theory and explicit forward simulations} to gain a better understanding of how this impacts patterns of introgression \yb{across genomic and geographic} space. \alisa{We describe the properties of clines across the genomes as they are formed, as well as correlations in ancestry. Clines are expected to form over time $1/s$, where $s$ is the strength of selection against hybrids. Linked clines persist over a  physical scale of $1/T$, where $T$ is the age of the hybrid zone. Additionally, we find that locally disadvantageous alleles move at speed $\sigma\sqrt{s}$ (where $\sigma$ is the variance in dispersal), and tend to exist as relatively recent, small families. The length of continuous tracts of ancestry provide an additional source of information; within a population, blocks of ancestry surrounding single-locus incompatibilities} are longer than the genome-wide average block length, especially when removed from the center of the hybrid zone.  The findings presented here suggest a means for characterizing hybrid zones and identifying candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}

@@

%\plr{
%Main points, make sure they appear in the abstract:
%clines establish over time $1/s$, extend along genome to at least $1/T$;
%locally disadvantageous alleles exist as recent, small families (moving at speed $\sigma/\sqrt{s}$);
%haplotype chunks longest around introgressing selected bits, flanked by shorter chunks.
%}

@@

The process of speciation commonly involves allopatric divergence in relative isolation \citep{Coyne2004}.  
When formerly isolated populations come back into contact 
before reproductive isolation is complete, gene flow is still possible. 
As two such populations interbreed, tracts of
individuals' genomes can be labeled according to which ancestral population they are inherited from,
and local interbreeding and migration of such populations
create a gradient of ancestry assignments across geographic space,
centered on their point of contact \citep{Barton1985}. 
%\plr{(check that neutral clines discussed in barton '85)}
If the populations are sufficiently diverged,
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
leading to substantial correlations in ancestry between distant markers.
The influence of selection on linked loci also prevents the mixing of ancestry at regions surrounding these loci. %the effect of which decreases with physical genomic distance \citep{Barton1986,Barton1983}. 
The mixing of ancestries of a given genome is mediated by 
the process of recombination, forming chromosomal \emph{junctions} in ancestry \citep{Fisher1954, Chapman2002, Baird2003}.   
Patterns of correlated ancestry in admixed populations has been previously described under neutral conditions both in single populations and in contact zones \citep[e.g.][]{Gravel2012,Hellenthal2014,Sedghifar2015}.

While for much of the genome gene flow will eventually erase 
both geographic clines in ancestry, and strong correlations in ancestry,
subsets of the genome can maintain stable clines in ancestry if selection against migrant, or hybrid, genotypes impedes gene flow across the hybrid zone \citep{Barton1979a}. 
% Variation in \yb{ the action of } selection against migrant alleles within the genome results in genome-wide heterogeneity in cline width, as the frequency of foreign ancestry surrounding targets of selection is reduced \citep{Barton1979}. 
% While these patterns of co-ancestry have been described in multi-locus systems, the \yb{expected length of ancestral tracts around selected loci in admixing populations remains poorly understood}\alisa{, especially in the geographic context of hybrid zones.}
\yb{While empirical patterns of the block length distribution in hybrid zones have rarely been used to inform the history of selection,}  \alisa{the availability of dense genotype data in combination with linkage information pave a way to utilize such patterns to better understand the processes operating in hybrid zones. } %GREAT SENTENCE!!
Here, we investigate expected patterns of co-ancestry \yb{ --- that is, the genomic scale at which ancestry strongly covaries between sites --- } across a hybrid zone in the presence of selection.

The extensive history of population genetic modeling of hybrid zones has focused largely on the equilibrium states of ancestry clines \citep{Barton1979a,Barton1986}.  
Since stronger selection reduces the amount of genetic mixing,
these models predict the narrowest clines 
(i.e., the most rapid changes in allele frequencies)
at loci close to selected loci, and wider clines at less tightly linked loci.
This provides some expectation for how selection shapes covariance in ancestry at a single locus over \emph{geographic space}.
Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry over \emph{genomic space}.

Covariance in ancestry across a genome provides information about the number of surviving recombination events that have taken place between ancestral genotypes and therefore provides information about the age of the hybrid zone and the relative strength of selection affecting subsets of the genome. Our goal is to understand how selection against hybrids interacts with dispersal and recombination to produce observed patterns of ancestry around selected sites in recently established hybrid zones. In addition to interpreting non-equilibrium patterns of diversity in hybrid zones, this could be especially important for understanding the impact of selection on the extent of mixing of ancestral genomes and, consequently, the maintenance of species distinctness. In the extreme case of zero fitness in first-generation hybrids, for example, parental genomes will never mix and the stable hybrid zone will resemble a recently formed one. 
If, on the other hand, hybrids have non-zero fitness, some amount of introgression can occur, and providing information about the extent of hybridization and timing of secondary contact \citep{Price2009, Hellenthal2014}. 
\yb{In this latter case, heterogeneous patterns of ancestry across the genome potentially allows researchers to identify putative targets of selection in hybridizing populations.}  
\yb{To date, researchers have done so by examining heterogeneous clines in ancestry across loci \citep{Porter1997, Gompert2012}, but have not made use of the information contained in the length of ancestry blocks around putatively selected loci.}

Under a model of selection against hybrids, migrant haplotypes containing the selected locus will be removed rapidly from the population, preventing introgression in surrounding genomic regions. We therefore expect \yb{a deficit of} short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. As a corollary to this, conditional on having the ancestry that is at lower frequency (that is, being on the `wrong' side of the hybrid zone), the length of unbroken ancestry surrounding the selected locus is expected to be relatively long when far away from the zone center.  This is because an unfit haplotype is more likely to be a recent migrant from the other side of the hybrid zone, and therefore will not have experienced many recombination events.

Using a combination of theory and simulated hybrid zones, we present a model of genome-wide patterns coancestry as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center. For this, we build on the neutral framework presented in \citet{Sedghifar2015} to derive the expected distribution of ancestry block lengths across a contact zone.  
\yb{The theoretical expectations developed in this manuscript can aid in the identification of loci under selection in hybridizing populations.}

@@

We model selection through a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations.
Individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are very similar to models of ecological selection or epistatic systems \cite{Kruuk1999,Barton2000a}.

As in \citet{Sedghifar2015}, 
we gain insight by considering where the ancestors of modern-day individuals lived,
i.e., where their lineages fall across geography as we look further back towards the time of first contact.
We say that a locus in a sampled individual is of \emph{ancestry} $A$ if it has been inherited from an individual of species $A$ at the time of secondary contact,
i.e., if its lineage traces back to to the $x<0$.
A block of genome is of ancestry $A$ if every locus in it does the same;
recombination events may cause us to follow more than one lineage back to determine the block's ancestry,
and the block will be of ancestry $A$ if all lineages trace back to the $A$ side.
This process, in which recombination events cause such lineages to \emph{branch},
is illustrated in Figure~\ref{Fig:schematic}.
We say that a block of genome is on the $A$ \emph{background} if it is physically linked to an allele of ancestry $A$ at the selected locus;
if a block of ancestry $A$ includes the selected site then it is necessarily on the $A$ background.
As the identity of the selected allele determines how selection acts on the haplotype,
and linked variation can only move between the $A$ and the $B$ background in heterozygotes,
a key factor in these models is the density and fecundity of heterozygotes.

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citep{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, even in steady state, 
a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on numerous approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. 
%\alisa{I don't think I understand the independence from $\sigma$?}
%\plr{It gets set up at speed $\sigma$ over a distance proportional to $\sigma$.}

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele $p(x,t)$ is known.
We seek to first understand how lineages of sampled individuals behave at the selected locus,
from which lineages at linked loci require only a minor modification.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $\sigma^2 \grad \log p(x,t)$ (i.e., it moves at speed $\sigma^2$ in the potential $-\log p$).
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x+r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$.

At a geographic position far from the center, $p=1$ to the left, and is proportional to  $\exp(-x\sqrt{s}/\sigma)$ to the right.
Therefore, roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma\sqrt{s}$ 
back towards the side where they are more common. 
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact, this push must get more intense the closer it is to the time of secondary contact.
Lineages of $B$ alleles do the same thing, of course, but in the opposite direction. 
% \alisa{I want to make a figure from the simulations that show the rapid retreat of wrong ancestries back to where they came from}

@@

Suppose we sample an allele today at a locus at recombination distance $r$ to the selected site.
We will follow its lineage back through time, using $\tau$ to denote ``generations ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ (same as previously described movement of a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$)or $r p$ (for $B$ to $A$).

@@

@@

\paragraph{Numerical solutions}
Notice that equations \eqref{eqn:g_pde}
are hierarchical in $(a,b)$:
the equation for haplotype identity probabilities on a segment $(a,b)$ depends only on those probabilities for segments contained in $(a,b)$.
This allows for efficient numerical solutions,
described in more detail in Appendix \ref{apx:haplotype_calcs}.

@@

\begin{figure}
\includegraphics{figs/alleleFrequencies_sim}
\caption{
    Frequency of ancestry $B$,
    across geography at different physical positions on the genome, simulated for a hybrid zone
     $T=100$ generations after secondary contact,
    and with $s=0.1$.
        $50$ demes, each with population size $N_d=500$ diploid individuals, were simulated across the hybrid zone.
    Each line represents a locus some distance $r$ away from the true target of selection, 
with $r=0$ representing the locus under selection. Grey lines represent the same positions from a simulation with
    identical parameters except $s=0$.
    Corresponding theoretical quantities are shown juxtaposed in Figure \ref{sfig:alleleFreq_tau100_comparison}.
}\label{alleleFreq_tau100}
\end{figure}

@@

The above theory and simulations allow us to make predictions about patterns surrounding selected loci. In reality, however, such loci are not known, so it is useful to have per-site statistics that may allow for detection of candidate targets of selection. The most straightforward measure is $l_I(m)$,  the mean length of all haplotype chunks that contain genomic position $m$, given that $m$ is of ancestry $I$, where $I\in\{A,B\}$.
We then focus on $l_B(m)$ for $m$ across a chromosome. 
%Another statistic is the cumulative probability $C_I(m)$ of a sampled chunk encompassing position $m$ being longer than a chunk sampled from the distribution of all chunk lengths across the genome in the population.

@@

Although our analytical model does not account for coalescence, we are able to partially trace the genealogy of haplotypes in our simulations. In particular, we wish to know the relative number of ancestors from $T$ generations ago that are represented in present day populations at a given locus, given the frequency of a particular ancestry. This is a reflection of the average size of a haplotype family, and is potentially a source of additional information about selection. Within each deme $x$, we calculated $F_B(m,x)$, which is the number of independent ancestors from time $T$ contributing to the pool of alleles of ancestry $B$ at site $m$, divided by the number of individuals of ancestry $B$ at site $m$.
%\plr{how about $x$ instead of $X$? above $x$ refers to spatial location.}

@@

\paragraph{The timescale over which linked clines relax:}
Although loci not under direct selection can in principle spread across the cline unimpeded,
in practice it can take quite some time for even unlinked neutral loci to homogenize,
due to the decreased fitness of heterozygotes \citep{Barton1986}
and the relative slowness of diffusive movement \citep{Sedghifar2015}.
We are interested, for samples across a cline $T$ units of time after secondary contact, how far away from the selected locus one has to look to see clines that resemble the genome-wide mean.

@@

@@

We expect that, in the absence of selection, blocks of $A$ ancestry across the genome 
will tend to be shorter the further one goes into the $B$ side of the cline,
because they have had more opportunities to recombine with $B$ haplotypes. 
However, we expect that stretches of $A$ ancestry containing a selected site will be longer
than those that do not contain the selected site at the same spatial location,
because lineages containing the selected site have usually been inherited from the $A$ side of the cline recently. 
As discussed above, we expect these lineages move at speed roughly $\sigma\sqrt{s}$,
so (selected) $A$ alleles at distance $x$ from the cline center
to have last had an ancestor on the $A$ side of the cline around $x /\sqrt{s}\sigma$ generations ago 
(compared to $x^2/\sigma^2$ for a neutral allele),
and so the enclosing $A$ haplotype is no longer than $\sigma\sqrt{s}/x $.
%\plr{(check predicted decay of mean length $1/x$ in simulations?)}
For linked haplotypes not containing the selected site, the situation is more complicated, and harder to intuit.

@@

% This pattern provides a potential avenue for identifying targets of selection by looking for a signal of increased $l_B$ or $l_A$ for positions surrounding a selected site, compared to the genome-wide background  at geographic sites far from the center of the zone (Figure~\ref{Fig:blockLengths},~\ref{Supp:blockLengthHeatmap}). 
The mean haplotype length found without conditioning on ancestry ($l_\cdot(m)$)
also shows an small increase near the selected locus,
but it is much smaller (see Figures \ref{Supp:blockLengthHeatmapNoAnc} and \ref{Supp:blockLengthNoAnc}).

\alisa{For weaker selection ($s=0.001$), then the relative speed of retreat is not fast enough compared to the genomic background, and the strength of signal from elevated $l_B(m)$ is diminished. The resolution of the peak is expected to scale with $1/T$, as this is the physical scale over which linked clines persist. Since clines in ancestry establish over the timescale $1/s$, it is difficult to intuit the time it takes for a signal of elevated $l_B(m)$ to form around the selected locus.

We note that for very strong $s$, few haplotypes are found on the ``wrong" side, and so the signal of longer haplotypes around selected loci may be present only at geographic locations close to the center of the zone.}

%\alisa{For very strong $s$, very few haplotypes are found on the ``wrong" side, and so the signal of longer haplotypes around selected loci may only be present at geographic locations close to the center of the zone. The overall length of the haplotypes decreases as the hybrid zone ages, but the signal remains strong relative to the rest of the genome and detectability will largely depend on the density of ancestry-informative markers.}

@@

@@
